What Makes Us Happy? Examining the Impact of Social, Economic, and Personal Factors

### 2024-12-03 paper/paper.qmd
make some changes

The essential role of socio-economic stability in fostering happiness is well-documented. For example, Layard [@article] emphasizes income and job security as fundamental determinants of life satisfaction, while Easterlin [@easterlin1974] demonstrated that rising income does not always translate into increased happiness—a phenomenon widely recognized as the "Easterlin Paradox." More recently, Helliwell et al. [@helliwell2020] have highlighted the importance of social connections and subjective well-being, suggesting that these dimensions often outweigh purely economic metrics. Despite these valuable contributions, a notable issue persists in understanding how socio-economic factors interact with personal characteristics—such as marital status, gender, and the number of children—in shaping happiness.

@@

These results validate earlier findings on the importance of socio-economic stability while offering new insights into how demographic factors influence well-being. The study provides practical recommendations for policymakers and social scientists to design targeted interventions aimed at enhancing happiness across different population segments.

@@

In addition to the GSS dataset, there are other datasets that could have been used to analyze the factors influencing happiness. For example, the European Social Survey (ESS) provides rich data on life satisfaction and happiness, but it primarily focuses on European countries, making it less suitable for this study's focus on American society. In contrast, the GSS dataset is centered on the U.S. population, offering more relevant data for examining the impact of socio-economic conditions and personal characteristics on happiness in the United States. Similarly, the World Values Survey (WVS) offers global social-economic data, but due to its broad scope, it may not be as suitable for focusing on happiness factors specific to the U.S. population. The American Time Use Survey (ATUS) primarily focuses on time allocation rather than directly measuring life satisfaction or happiness, making it less aligned with the objectives of this study. The GSS dataset was chosen for its focus on U.S. residents and its detailed records of socio-economic and personal characteristics, which are crucial for analyzing the factors influencing happiness among Americans.

@@

- **`tidyverse`** [@tidyverse]: Provided a suite of tools for data manipulation and visualization, significantly simplifying the workflow.
- **`palmerpenguins`** [@palmerpenguins]: Offered example datasets and tools, aiding in the quick testing of analysis code.
- **`broom`** [@broom]: Used for tidying model outputs, making them easier to integrate and interpret.
- **`ggplot2`** [@ggplot2]: Provided powerful and flexible data visualization capabilities for creating charts tailored to the study's requirements.
- **`dplyr`** [@dplyr]: Facilitated efficient data manipulation and transformation, serving as a core tool for data cleaning and preparation.
- **`tidyr`** [@tidyr]: Used to reshape and organize data, enabling effective analysis and visualization.
- **`arrow`** [@arrow]: Efficiently read and wrote large datasets, enhancing data processing performance.
- **`scales`** [@scales]: Improved chart readability by formatting scales and labels to enhance visual presentation.
- **`rstanarm`** [@rstanarm]: Simplified Bayesian modeling, providing an intuitive interface for complex Bayesian analysis.
- **`brms`** [@brms]: A flexible modeling tool built on Stan, used to perform Bayesian regression analysis on the data.
- **`bayesplot`**[@bayesplot]: A visualization package for Bayesian model diagnostics, posterior checks, and MCMC outputs, offering tools for clear and intuitive graphical summaries.

@@

**`age`**: Represents the respondent's age, measured in years.

**`sex`**: Denotes the biological sex of the respondent, categorized as: **male** and **female**.

**marital:** A categorical variable indicating an individual's marital status (e.g., "Married"). "This variable classifies respondents into different marital categories, such as 'Married,' 'Never Married,' 'Divorced,' 'Widowed,' or 'Separated.' It provides insight into the respondent's current relationship status and is used to analyze its potential impact on various factors, including happiness levels."

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-ChildrenDistribution
#| fig-cap: "Distribution of the Number of Children by Happiness Level: Showing Family Size Across Different Happiness Categories"

# Filter out non-finite values (NA or Inf)
data <- data %>%
  filter(!is.na(childs), is.finite(childs))  # Remove NA or infinite values

# Create a violin plot with jittered points
ggplot(data, aes(x = happy, y = childs, fill = happy)) +
  geom_violin(trim = TRUE, alpha = 0.7) +  # Violin plot to show distribution
  geom_jitter(width = 0.2, alpha = 0.4, color = "black", size = 1) +  # Add jitter points for raw data
  scale_fill_manual(values = c(
    "pretty happy" = "#41AB5D",  # Dark green
    "very happy" = "#00441B",    # Medium green
    "not too happy" = "#A1D99B"  # Light green
  )) +
  theme_minimal() +
  labs(
    x = "Happiness Level",
    y = "Number of Children",
    fill = "Happiness Level"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    plot.title = element_text(face = "bold", hjust = 0.5),  # Bold and center the title
    legend.title = element_text(face = "bold"),  # Bold legend title
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()   # Remove minor gridlines
  )

```
@fig-ChildrenDistribution illustrates the relationship between the number of children and happiness levels, with data points grouped by happiness categories for better clarity. The violin plots display the density distribution of the number of children across the three happiness levels: "Not Too Happy," "Pretty Happy," and "Very Happy," differentiated by distinct shades of green. Additionally, jittered points overlay the violin plots, representing individual observations, thereby providing a detailed view of the raw data.

The chart highlights that the number of children is distributed similarly across happiness levels, with the highest density of observations clustering around smaller family sizes. A few data points stand out as outliers, representing families with larger numbers of children. This combination of violin plots and jittered points effectively showcases both the overall trends and individual variances, making the relationships between family size and happiness levels visually accessible and easy to interpret.

@@

Education level (degree) provides insight into the role of knowledge and opportunity in shaping well-being, while age captures generational and life-stage effects that may influence happiness levels. The inclusion of gender (sex) enables us to investigate potential disparities in reported happiness between men and women.

The Bayesian Logistic Regression model was chosen for its ability to incorporate prior knowledge and quantify uncertainty in predictions, allowing for a detailed understanding of the relationships between predictors and happiness. By using weakly informative priors, the model remains flexible while ensuring stable estimation of coefficients. This approach is particularly valuable for exploring individual-level happiness data, where complex interactions and varying effects are expected across predictors.

@@

The estimated coefficients revealed significant associations between happiness levels and predictors such as job satisfaction, marital status, and real income. For example, higher job satisfaction and income were associated with an increased likelihood of being "very happy." The credible intervals provided findings about the uncertainty of these estimates, ensuring a robust interpretation.

@@

## Model Limitations

Although the Bayesian logistic regression model performs well in analyzing the relationship between happiness and its predictors, it still has certain limitations. First, the model assumes a linear relationship between the predictors and the log-odds of happiness, which may overlook complex non-linear or interaction effects between variables. For example, the relationship between age and happiness may exhibit non-linear patterns, leading to potential biases in estimation. Second, the imbalanced distribution of categories in the data—where the "very happy" category has fewer samples—may cause the model to favor predicting the majority classes (e.g., "not too happy" and "pretty happy") while underperforming in classifying the minority class. Lastly, the model employs weakly informative priors (e.g., Normal(0, 2) and Cauchy(0, 2)) to ensure flexibility. However, in cases with small samples or rare categories, these priors may impact the stability of coefficient estimates, increasing uncertainty in the results.

@@

The Bayesian logistic regression model was chosen due to its unique ability to incorporate prior knowledge, quantify uncertainty, and provide robust parameter estimates in the analysis of complex individual-level happiness data. Unlike traditional logistic regression, the Bayesian framework allows for the inclusion of weakly informative priors, which help stabilize estimates in the presence of limited or imbalanced data (e.g., the "very happy" category being underrepresented) while avoiding overfitting. This is particularly advantageous in handling imbalanced target variable distributions, where the Bayesian model performs better in classifying minority categories. Additionally, the Bayesian model computes posterior distributions for each predictor, offering a comprehensive understanding of the uncertainty and variability in the estimated effects. This probabilistic framework not only enhances interpretability but also supports better decision-making by providing credible intervals instead of single-point estimates.

@@

For these reasons, the Bayesian logistic regression model stands out for its flexibility and robustness, capturing complex non-linear relationships and potential interactions while quantifying uncertainty. It offers more accurate and comprehensive insights for predicting happiness compared to alternative models, making it a superior choice for this analysis.

@@

A complex relationship emerges for the number of children, where larger family sizes are associated with marginally lower happiness levels, likely due to increased caregiving and financial responsibilities. Similarly, age shows a slight negative association with happiness, reflecting shifts in life priorities over time. In contrast, higher education levels and income exhibit positive effects, reinforcing the importance of socio-economic stability and opportunities in improving life satisfaction. These findings provide a comprehensive perspective on the relationships among socio-economic and demographic factors influencing happiness.

@@

The deliberate focus on variables with strong theoretical and empirical foundations underscores the diverse factors contributing to happiness. Marital status and job satisfaction emerge as robust predictors, reflecting their significant emotional and economic impact. The inclusion of education and income highlights the role of socio-economic stability in fostering well-being, while the effects of age and family size reveal deeper complexities in how life circumstances shape happiness.

Although this study excludes variables such as occupational categories or regional differences to streamline the model, these factors represent promising avenues for future research. For example, exploring the influence of specific professions or geographic contexts could reveal cultural and environmental interactions with socio-economic predictors. Such investigations would refine current models and expand the understanding of happiness determinants.

@@

Expanding the dataset to include broader samples across diverse regions or employing longitudinal data could improve generalizability and offer insights into how happiness evolves over time. Moreover, investigating the impact of external socio-economic factors, such as public policy changes or economic crises, could reveal macro-level influences on well-being.

Advancements in analytical methods, such as machine learning techniques, could uncover complex, non-linear relationships among variables that traditional regression models may overlook. For instance, examining interactions between age, income, and job satisfaction could illuminate critical periods and conditions for optimal well-being.

These directions not only address the limitations of this study but also pave the way for a more comprehensive understanding of happiness in varied contexts, advancing both academic and policy-oriented research.

@@

## The Value of Insights into Happiness Drivers

@@

## Summary
These cleaning steps were essential for maintaining data integrity and ensuring that the analysis would provide meaningful and reliable results. By standardizing and filtering the data, the study could focus on valid and interpretable responses, minimizing noise and maximizing the quality of insights derived from the dataset.

@@

Our goal is to provide a robust analysis of the factors influencing individual happiness by designing a high-quality survey that captures data on key socio-economic and demographic variables. By addressing potential biases and ensuring representative sampling, this methodology aims to produce actionable insights within a manageable budget.

@@

This comprehensive methodology ensures reliable insights into the complex factors shaping individual happiness while laying the groundwork for future studies on well-being determinants.

@@

Welcome to the "Factors Influencing Happiness Survey." This survey seeks to explore the relationship between socio-economic factors, personal characteristics, and individual well-being. Your responses will remain anonymous and contribute to research aimed at understanding the determinants of happiness. Completing the survey will take approximately 5–7 minutes. Thank you for your valuable participation!

>>>>

The essential role of socio-economic stability in fostering happiness is well-documented. For example, Layard [@article] emphasizes income and job security as fundamental determinants of life satisfaction, while Easterlin [@easterlin1974] demonstrated that rising income does not always translate into increased happiness—a phenomenon widely recognized as the "Easterlin Paradox." More recently, Helliwell et al. [@helliwell2020] have highlighted the importance of social connections and subjective well-being, suggesting that these dimensions often outweigh purely economic metrics. Despite these meaningful contributions, a notable issue persists in understanding how socio-economic factors interact with personal characteristics—such as marital status, gender, and the number of children—in shaping happiness.

@@

These results validate earlier findings on the importance of socio-economic stability while offering new findings into how demographic factors influence well-being. The study provides practical recommendations for policymakers and social scientists to design targeted interventions aimed at enhancing happiness across different population segments.

@@

In addition to the GSS dataset, there are other datasets that could have been used to analyze the factors influencing happiness. For example, the European Social Survey (ESS) provides rich data on life satisfaction and happiness, but it primarily focuses on European countries, making it less suitable for this study's focus on American society. In contrast, the GSS dataset is centered on the U.S. population, offering more relevant data for examining the impact of socio-economic conditions and personal characteristics on happiness in the United States. Similarly, the World Values Survey (WVS) offers global social-economic data, but due to its broad scope, it may not be as suitable for focusing on happiness factors specific to the U.S. population. The American Time Use Survey (ATUS) primarily focuses on time allocation rather than directly measuring life satisfaction or happiness, making it less aligned with the objectives of this study. The GSS dataset was chosen for its focus on U.S. residents and its detailed records of socio-economic and personal characteristics, which are essential for analyzing the factors influencing happiness among Americans.

@@

- **`tidyverse`** [@tidyverse]: Provided a suite of tools for data manipulation and visualization, significantly simplifying the workflow.
- **`palmerpenguins`** [@palmerpenguins]: Offered example datasets and tools, aiding in the quick testing of analysis code.
- **`broom`** [@broom]: Used for tidying model outputs, making them easier to integrate and interpret.
- **`ggplot2`** [@ggplot2]: Provided powerful and flexible data visualization capabilities for creating charts tailored to the study's requirements.
- **`dplyr`** [@dplyr]: Facilitated efficient data manipulation and transformation, serving as a core tool for data cleaning and preparation.
- **`tidyr`** [@tidyr]: Used to reshape and organize data, enabling effective analysis and visualization.
- **`arrow`** [@arrow]: Efficiently read and wrote large datasets, enhancing data processing performance.
- **`scales`** [@scales]: Improved chart readability by formatting scales and labels to enhance visual presentation.
- **`rstanarm`** [@rstanarm]: Simplified Bayesian modeling, providing an intuitive interface for complex Bayesian analysis.
- **`brms`** [@brms]: A flexible modeling tool built on Stan, used to perform Bayesian regression analysis on the data.
- **`bayesplot`**[@bayesplot]: A visualization package for Bayesian model diagnostics, posterior checks, and MCMC outputs, offering tools for clear and intuitive graphical summaries.
- **`testthat`** [@testthat]: Enabled modular and reproducible testing of data and analysis workflows, ensuring the validity of datasets, functions, and model outputs while enhancing code reliability.

@@

**age**: Represents the respondent's age, measured in years.

**sex**: Denotes the biological sex of the respondent, categorized as: male and female.

**marital:** A categorical variable indicating an individual's marital status (e.g., "Married"). "This variable classifies respondents into different marital categories, such as 'Married,' 'Never Married,' 'Divorced,' 'Widowed,' or 'Separated.' It provides finding into the respondent's current relationship status and is used to analyze its potential impact on various factors, including happiness levels."

@@

@@

Education level (degree) provides finding into the role of knowledge and opportunity in shaping well-being, while age captures generational and life-stage effects that may influence happiness levels. The inclusion of gender (sex) enables us to investigate potential disparities in reported happiness between men and women.

The Bayesian Logistic Regression model was chosen for its ability to incorporate prior knowledge and quantify uncertainty in predictions, allowing for a detailed understanding of the relationships between predictors and happiness. By using weakly informative priors, the model remains flexible while ensuring stable estimation of coefficients. This approach is particularly meaningful for exploring individual-level happiness data, where complex interactions and varying effects are expected across predictors.

@@

The estimated coefficients indicated significant associations between happiness levels and predictors such as job satisfaction, marital status, and real income. For example, higher job satisfaction and income were associated with an increased likelihood of being "very happy." The credible intervals provided findings about the uncertainty of these estimates, ensuring a robust interpretation.

@@

## Model Assumptions and Limitations

The Bayesian logistic regression model is built upon several key assumptions that underpin its structure and interpretation. First, the model assumes a linear relationship between the predictors and the log-odds of happiness, meaning that the effects of predictors remain constant across their range. This assumption simplifies the modeling process but may fail to capture complex non-linear patterns or interactions between variables, such as potential non-linear relationships between age and happiness. The model also assumes independence between predictors, which may not hold in real-world data where multicollinearity or interaction effects are common. Additionally, the model relies on weakly informative priors (e.g., Normal(0, 2) for coefficients and Cauchy(0, 2) for the intercept) to provide flexibility while avoiding overfitting. While these priors stabilize the estimation process, they may still introduce bias or increase uncertainty in cases with small samples or sparse categories. Lastly, the model assumes homoscedasticity and Gaussian-distributed residuals, which may not fully account for variability or skewness in happiness data.

Despite its strengths, the model has certain limitations that should be noted. First, the linearity assumption between predictors and log-odds may oversimplify relationships, leading to potential biases in effect estimation. For instance, the relationship between income and happiness might exhibit diminishing returns, which a linear model may struggle to capture. Second, the imbalanced distribution of the target variable—where "very happy" is a minority category—presents a challenge, as the model may prioritize predicting majority categories ("not too happy" and "pretty happy") while underperforming in classifying the minority category. Additionally, while weakly informative priors are generally robust, they may lead to unstable coefficient estimates in cases of sparse data or rare categories, increasing uncertainty in the results. The assumption of homoscedasticity may also be violated in situations where variability in happiness levels depends on specific predictors, such as income or job satisfaction.

@@

The Bayesian logistic regression model was chosen due to its unique ability to incorporate prior knowledge, quantify uncertainty, and provide robust parameter estimates in the analysis of complex individual-level happiness data. Unlike traditional logistic regression, the Bayesian framework allows for the inclusion of weakly informative priors, which help stabilize estimates in the presence of limited or imbalanced data (e.g., the "very happy" category being underrepresented) while avoiding overfitting. This is particularly advantageous in handling imbalanced target variable distributions, where the Bayesian model performs better in classifying minority categories. Additionally, the Bayesian model computes posterior distributions for each predictor, offering a detailed understanding of the uncertainty and variability in the estimated effects. This probabilistic framework not only enhances interpretability but also supports better decision-making by providing credible intervals instead of single-point estimates.

@@

For these reasons, the Bayesian logistic regression model stands out for its flexibility and robustness, capturing complex non-linear relationships and potential interactions while quantifying uncertainty. It offers more accurate and detailed findings for predicting happiness compared to alternative models, making it a superior choice for this analysis.

@@

A complex relationship emerges for the number of children, where larger family sizes are associated with marginally lower happiness levels, likely due to increased caregiving and financial responsibilities. Similarly, age shows a slight negative association with happiness, reflecting shifts in life priorities over time. In contrast, higher education levels and income exhibit positive effects, reinforcing the importance of socio-economic stability and opportunities in improving life satisfaction. These findings provide a detailed perspective on the relationships among socio-economic and demographic factors influencing happiness.

@@

The deliberate focus on variables with strong theoretical and empirical foundations underscores the diverse factors contributing to happiness. Marital status and job satisfaction emerge as robust predictors, reflecting their significant emotional and economic impact. The inclusion of education and income highlights the role of socio-economic stability in fostering well-being, while the effects of age and family size indicate deeper complexities in how life circumstances shape happiness.

Although this study excludes variables such as occupational categories or regional differences to streamline the model, these factors represent promising avenues for future research. For example, exploring the influence of specific professions or geographic contexts could indicate cultural and environmental interactions with socio-economic predictors. Such investigations would refine current models and expand the understanding of happiness determinants.

@@

Expanding the dataset to include broader samples across diverse regions or employing longitudinal data could improve generalizability and offer findings into how happiness evolves over time. Moreover, investigating the impact of external socio-economic factors, such as public policy changes or economic crises, could indicate macro-level influences on well-being.

Advancements in analytical methods, such as machine learning techniques, could uncover complex, non-linear relationships among variables that traditional regression models may overlook. For instance, examining interactions between age, income, and job satisfaction could illuminate essential periods and conditions for optimal well-being.

These directions not only address the limitations of this study but also pave the way for a more detailed understanding of happiness in varied contexts, advancing both academic and policy-oriented research.

@@

## The Value of Findings into Happiness Drivers

@@

## Summary
These cleaning steps were essential for maintaining data integrity and ensuring that the analysis would provide meaningful and reliable results. By standardizing and filtering the data, the study could focus on valid and interpretable responses, minimizing noise and maximizing the quality of findings derived from the dataset.

@@

Our goal is to provide a robust analysis of the factors influencing individual happiness by designing a high-quality survey that captures data on key socio-economic and demographic variables. By addressing potential biases and ensuring representative sampling, this methodology aims to produce actionable findings within a manageable budget.

@@

This detailed methodology ensures reliable findings into the complex factors shaping individual happiness while laying the groundwork for future studies on well-being determinants.

@@

Welcome to the "Factors Influencing Happiness Survey." This survey seeks to explore the relationship between socio-economic factors, personal characteristics, and individual well-being. Your responses will remain anonymous and contribute to research aimed at understanding the determinants of happiness. Completing the survey will take approximately 5–7 minutes. Thank you for your helpful participation!
